Data Provenance for Concurrent Program Debugging and Analysis in Production Systems

### 2013-11-11 paper.tex
committing new eval stuff

Understanding the connection between {\tt append()} and {\tt erase()} is the
key to fixing this bug.  The provenance of the value stored in {\tt sb->count}
at {\bf B} was the write at {\bf A} in {\tt erase()} and with that provenance
information, the programmer would understand that key connection.  Prior
provenance work on origin tracking for unusable values~\cite{badapples} does
not help here.  The involved values are inconsistent, but they are all {\em
usable} so their provenance would not be tracked.

Another challenge posed by this bug is that it manifests as a failure
infrequently and under only real-world conditions.  If the programmer runs the
program in a debugger, the failure may not manifest in a reasonable amount of
time and running it may require unavailable real-world inputs.  Thus,
collecting provenance or other information offline may be time-consuming or
impossible.  A better strategy is to continuously collect that information in
deployed systems and package it with core dumps sent to programmers with bug
reports.  In this work, we propose a last writer slicing design that espouses
this {\em in situ} data collection strategy and does so with overheads
appropriate for production systems.

\subsection{Tracking Dependences and Communication}
\label{sec:background:comm}
{\bf TODO: Clean this up, downplay the "dependences" obsession and make it 
a better match to the LWS mission, as stated in the intro -- provenance + instrumentation = analysis}
Monitoring inter-thread dependences is the foundation of many useful
concurrency analyses.  For example, in a software transactional memory (STM)
system~\cite{stm} dependences correspond to interference between accesses in
different threads that must be detected to rollback conflicting transactions.
Deterministic concurrent runtime systems~\cite{coredet,grace} are similar --
they monitor inter-thread dependences, to determine when to apply
deterministic ordering constraints on threads' accesses.  STMs and
deterministic systems must track all dependences and they must track them
precisely, to provide strong guarantees as programming and execution models.
Unfortunately, precision has a high runtime overhead -- {\em e.g.},
CoreDet incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is
too high for use in production.

%The overhead comes from the need to precisely track read-after-write (RAW),
%write-after-read (WAR), and write-after-write (WAW) dependences, taking action
%({\em e.g.}, buffering) on most cross-thread dependences.  These systems track
%dependences with additional code that runs around memory accesses.  To ensure
%correctness, accesses and tracking code usually must be protected by
%synchronization.  The high cost of synchronization exacerbates the overheads.

In contrast to {\em imprecision intolerant} tasks like STM and deterministic
execution, other tasks are more {\em imprecision tolerant}.  Imprecision
tolerant systems help perform tasks like
debugging~\cite{defuse,conseq,recon,bugaboo,raceslicing,fasttrack,falcon},
software engineering~\cite{oshajava,oshatr}, anomaly
detection~\cite{avio,dmtracker,cci,daikon}, bug and failure
avoidance~\cite{aviso,cfix} and concurrent performance
profiling~\cite{threadcriticality,schedpredictionmodel}.  These
systems also work by monitoring how data flows between threads.  They can
tolerate two important sources of imprecision.

First, many of these techniques track only an {\em ad hoc} subset of
dependences.  For example, DefUse~\cite{defuse} tracks inter-thread WAR
dependences, but not WAW dependences.  Recon~\cite{recon} tracks some inter-thread WAW and
RAW dependences and approximates WAR dependences.  These techniques are useful
despite an incomplete set of dependences.  Second, many of these systems rely
on statistical models of software
behavior~\cite{avio,cci,defuse,recon,bugaboo,falcon,dmtracker,aviso,threadcriticality,schedpredictionmodel}.
Models often tolerate noise, so many of these techniques can relax the
correctness constraints of their implementation and remain useful.  For
example, some systems have chosen not to enforce atomicity of each memory
access and its dependence tracking code, compromising precision for a boost in
performance.

These tools and techniques work in a variety of environments.  A common theme
is the advantage reaped from use in a deployment environment.  Debugging tools,
profilers, and anomaly detectors benefit from seeing diverse, real-world
behavior.  Dynamic failure avoidance techniques must work in deployed systems
to be effective.  Deployment environments demand low overheads.

The design for \ctraps that we develop in this work monitors a subset of
inter-thread dependences that correspond to inter-thread communication.
\ctraps provides enough precision to build useful analyses and runtime systems
with overhead low enough for use in production.

@@

At any point in an execution, the information stored in a memory location's
\lwt entry represents the provenance of the data stored in that memory
location.   Last writer slices do not explicitly record any information about
read operations and they do not need to to be useful.  Interrupting the
execution while debugging and examining an entry in the \lwt reveals the
provenance of a value just when it is needed.  For example, if an assertion
failure occurs due to an erroneous inter-thread data flow, the provenance of
the data in the assertion's condition helps explain why the failure occurred.
Computing last writer slices is fast enough to deploy and use continuously
(see Section~\ref{sec:eval}) and the \lwt can be saved with a core dump.
Thus, last writer slices can be made part of the standard debugging
information provided when a program fails.

@@

{\bf TODO: Update the text to explain how the last writer slice shown makes the conection between erase() and append(), showing the programmer their mistake.}

@@

Another important trait of last writer slices illustrated by
Figure~\ref{fig:httlws} is that last writer slices are reported back to the
programmer {\em with the core dump on a failure}.  Capturing good debugging
information is especially important for concurrency bugs that manifest very
rarely or in the presence of only peculiar real-world inputs.  A primary goal
of this work is to show that we can collect last writer slices with overheads
that are low enough for production use so they can be included pervasively in
bug reports.

@@

We illustrate the debugging benefits of last writer slices by using our system
to debug several real-world programs studied previously in the debugging
literature.

@@

\paragraph{HTTrack}
{\bf TODO: Deal with this duplication from the copy/paste}
Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
plays in debugging, using an example of a real bug from HTTrack, a web caching
tool.  The figure depicts two threads manipulating a data structure that
contains a mutex\_lock.  The program has an ordering violation bug: the
program crashes when Thread 2 tries to acquire the lock before Thread 1 has
initialized it.  Assuming the run time environment supports it, a core dump
delivered to the programmer after such a failure reveals very little about how
to fix the bug.  The lock is uninitizalized, so the core dump may show that it
holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
insight into the root cause of the failure, which is that the code permits a
use of the lock before its initialization.   Comparing the core dump from the
failing run to the memory image from a correct execution -- a helpful debugging
strategy -- reveals only that in a correct execution, the lock contains valid
mutex state.  This revalation is of limited use as it does not rule out
arbitrary data corruption as the bug's root cause.

@@

The vital, missing piece to solving this bug is provenance information.
Debugging would be easier if the core dump included information about the
thread and code point that last wrote the lock.  The programmer would see that
no thread had initialized the lock in the failing execution, but that Thread 1
initialized the lock in non-failing executions.  This information guides the
programmer to fix the bug by adding synchronization that ensures the correct
operation ordering.

Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  Recall
that the the program crashes at point {\bf B} when it calls {\tt
hts\_mutex\_lock()} with an invalid mutex.    The root cause of this bug is
that Thread 1 fails to initialize the mutex before Thread 2 uses that mutex.
The last writer slices in the figure make this root cause clear: in the
failing execution, the last writer slice indicates the mutex is uninitialized; in
the non-failing execution, the last writer slice indicates the mutex was 
initialized by Thread 1 at point {\bf A}.  This information tells the
programmer that there is an ordering relationship between {\bf A} and {\bf B}
that must be enforced to fix the bug.  Note that without last writer slices,
the program crashes at {\bf B} and the core dump provides no information about
the connection between point {\bf A} and point {\bf B} -- the programmer would be left to guess that the bug was an initialization problem.

@@

Bad Apple Tracking~\cite{bondapple} is also a form of debugging with data
provenance information. This work records the {\em origin} of an unusable value
and can report it to a programmer when it is used in a way that causes a
failure.

>>>>

Understanding the connection between {\tt append()} and {\tt erase()} is the
key to fixing this bug.  The provenance of the value stored in {\tt sb->count}
at {\bf B} was the write at {\bf A} in {\tt erase()}.  Provenance information
shows the programmer that key connection.  Prior provenance work on origin
tracking for unusable values~\cite{badapples} does not help here.  The involved
values are inconsistent, but they are all {\em usable} so their provenance
would not be tracked.

Another challenge posed by this bug is that it manifests as a failure
infrequently and under only real-world conditions.  Prior slicing and
provenance debugging
techniques~\cite{tipslicingsurvey,thinslicing,whylineicse}, have overheads too
high to use in production.  Used offline, these techniques may not observe the
failure in a reasonable amount of time and triggering the bug may require
unavailable real-world inputs making offline analysis impossible.   A better
strategy is to continuously collect that information in deployed systems and
package it with core dumps sent to programmers with bug reports.  In this work,
we propose a last writer slicing design that espouses this {\em in situ} data
collection strategy and does so with overheads appropriate for production
systems.

\subsection{Communication Tracking in Concurrency Analyses}
\label{sec:background:comm}

Many useful concurrency analyses rely on monitoring inter-thread
communication.  The domains of these analyses include 
debugging~\cite{defuse,conseq,recon,bugaboo,raceslicing,fasttrack,falcon},
software engineering~\cite{oshajava,oshatr}, anomaly
detection~\cite{avio,dmtracker,cci,daikon}, bug and failure
avoidance~\cite{aviso,cfix} and concurrent performance
profiling~\cite{threadcriticality,schedpredictionmodel}.  The breadth of
applications illustrates the generality of communication tracking.

Despite the potential of a general communication tracking mechanism, prior work
has largely relied on {\em ad hoc} communication analyses.  For example,
DefUse~\cite{defuse} tracks communication via read-after-write (RAW)
dependences, but not write-after-write (WAW) dependences.  Recon~\cite{recon}
tracks some inter-thread WAW and RAW dependences.  These analyses could both be
built with a general communication tracking mechanism.  Unfortunately, their
single-purpose implementations track slightly different information, making
them incompatible and difficult to compose.

These analyses vary in their purpose and environments, but a cross-cutting
theme is the advantage reaped from their use in a deployment environment.
Debugging tools, profilers, and anomaly detectors benefit from seeing diverse,
real-world behavior.  Dynamic failure avoidance techniques must work in
deployed systems to be effective.  Deployment environments demand low overheads
and all of these analyses benefit from a general, high-performance
communication tracking framework.  Unfortunately, many of these techniques are implemented using heavy-weight infrastructure, like binary instrumentation~\cite{pin}, making them inappropriate for use in production.

There is a need for a general communication tracking framework that has
performance overheads that are low enough to use in production systems.  As we
describe in Section~\ref{sec:ctraps}, we aim to satisfy that need in this work.

%For example, in a software transactional memory (STM) system~\cite{stm}
%communication corresponds to interference between accesses in different threads
%that must be detected to rollback conflicting transactions.  Deterministic
%concurrent runtime systems~\cite{coredet,grace} similarly monitor
%communication, to apply deterministic ordering constraints.  STMs and
%deterministic systems must track all dependences and they must track them
%precisely, to provide strong guarantees as programming and execution models.
%Unfortunately, precision has a high runtime overhead -- {\em e.g.}, CoreDet
%incurs a mean overhead of around 5x~\cite{coredet}.  This overhead is too high
%for use in production.

@@

At each point in an execution, a memory location's \lwt entry represents the
provenance of the data stored in that memory location.   Last writer slicing
performs no action on read operations and to be useful, it does not need to.
Instead, a programmer can examine an \lwt entry at any arbitrary point in an
execution ({\em e.g.}, at a breakpoint), including after a failure, and see a
value's provenance.  Including the \lwt with core dumps means provenance is
available when post-mortem debugging production failures.

Figure~\ref{fig:jdklws} illustrates how the information in the \lwt helps debug
the JDK stringbuffer bug introduced in Figure~\ref{fig:coreDumpFail}.  Recall
that the failure occurs because Thread 2's execution of {\tt erase()} violates
the atomicity of Thread 1's execution of {\tt append()}.  The \lwt entry for
{\tt sb->count} at {\bf B} shows the programmer that its last writer was Thread
2 at {\bf A}. Thread 2 may have run ahead to an arbitrary point in its
execution.  Nothing in the core dump implicates {\tt erase()} in causing
this bug.  The last writer slice reveals the interference between {\tt append()}
and {\tt erase()} via {\tt sb->count}, helping the programmer understand the
bug's cause.

@@

@@

Another important trait of last writer slices illustrated by
Figure~\ref{fig:jdklws} is that when the assertion fails at {\bf B}, stopping
the execution, the \lwt is saved and can be sent to the programmer with the
core dump.  Capturing good debugging information is especially important for
concurrency bugs that manifest very rarely or in the presence of only peculiar
real-world inputs.  A primary goal of this work is to show that we can collect
last writer slices with overheads that are low enough for production use so
they can be included pervasively in bug reports.

@@

We illustrate the debugging benefits of last writer slices by using our system
to debug several real-world programs studied previously in the debugging
literature.  Table~\ref{tab:bugs} provides an overview of the bugs we studied
describing the program, bug type, and bug report identifier where applicable.
We then provide a summary of last writer slicing's debugging benefits.  To do
so, we mirror the evaluation strategy of bad value origin
tracking~\cite{badapples}, as it is the only other low-overhead provenance
tracking mechanism we know of.  We evaluate our system using their three
evaluation criteria - (1) effectiveness, (2) triviality, and (3) usefulness.
Our technique is effective if the last writer slice reports a code point that
wrote a value that subsequently caused a failure.  A failure is trivially
debuggable if its cause is made obvious by examining the core dump alone, {\em
i.e.}, a last writer slice is unnecessary.  A last writer slice is useful if it
is effective and the thread and code point in the slice actually help
understand why the failure occurred or how to fix it.   To draw contrast to bad
value origin tracking, we indicate whether their analysis could help debug each
bug, based on the description of their technique.

\begin{table}
\begin{tabular}{l|l|ccc}
{\bf App} & {\bf Type} & {\bf Eff.?} & {\bf Triv.?} & {\bf Usef.?} \\ \hline
mysql   & Atom.  &Yes  &No & Yes    \\
pbzip2  & Order  &Yes  &No & Yes    \\
httrack & Order  &Yes  &No & Yes    \\
jdk-sb  & Atom.  &Yes  &No & Yes    \\
apache1 & Atom.  &Yes  &No & Yes    \\ 
apache2 & Atom.  &Yes  &No & Maybe  \\
\end{tabular}
\caption{{\bf Evaluation of debugging with last writer slices.} We describe each bug and summarize the debugging benefit of last writer slices using the evaluation criteria from ~\cite{badapples}.}
\end{table}

@@

\paragraph{HTTrack}
%{\bf TODO: Deal with this duplication from the copy/paste}
%Figure~\ref{fig:coreDumpFail} illustrates the role that provenance information
%plays in debugging, using an example of a real bug from HTTrack, a web caching
%tool.  The figure depicts two threads manipulating a data structure that
%contains a mutex\_lock.  The program has an ordering violation bug: the
%program crashes when Thread 2 tries to acquire the lock before Thread 1 has
%initialized it.  Assuming the run time environment supports it, a core dump
%delivered to the programmer after such a failure reveals very little about how
%to fix the bug.  The lock is uninitizalized, so the core dump may show that it
%holds {\tt NULL} or it may hold random bits.  Both scenarios provide no
%insight into the root cause of the failure, which is that the code permits a
%use of the lock before its initialization.     This revalation is of limited use as it does not rule out
%arbitrary data corruption as the bug's root cause.
%
%\begin{figure}[h]
%\centering
%\includegraphics[width=\columnwidth]{figs/LWSHTTDebug.pdf}
%\caption{\label{fig:httlws}{\bf Using last writer slices to debug an
%ordering violation in HTTrack.} The provenance information stored in the last
%writer slice for the mutex variable helps understand the bug's root cause.}
%\end{figure}
%
%
%The vital, missing piece to solving this bug is provenance information.
%Debugging would be easier if the core dump included information about the
%thread and code point that last wrote the lock.  The programmer would see that
%no thread had initialized the lock in the failing execution, but that Thread 1
%initialized the lock in non-failing executions.  This information guides the
%programmer to fix the bug by adding synchronization that ensures the correct
%operation ordering.

Figure~\ref{fig:httlws} shows how last writer slices help understand the root
cause of the concurrency bug from Figure~\ref{fig:coreDumpFail}.  The program
crashes at point {\bf B} when it calls {\tt hts\_mutex\_lock()} with an invalid
mutex.    The root cause of this bug is that Thread 1 fails to initialize the
mutex before Thread 2 uses it.   The core dump is of little debugging use, only
showing that, indeed, the lock is uninitialized.  Even if the programmer
compared the core dump from the failing run to the memory image from a correct
execution -- a helpful debugging strategy -- they would find only that in a
correct execution, the lock contains valid mutex state.

@@

Looking at last writer slices make this root cause clear: in the failing
execution, the last writer slice indicates the mutex is uninitialized; in the
non-failing execution, the last writer slice indicates the mutex was
initialized by Thread 1 at point {\bf A}. While the core dump provides no
information about the connection between point {\bf A} and point {\bf B}, the
last writer slice reveals that the programmer should ensure {\bf B} follows
{\bf A} to prevent the failure.  Note that even augmenting the core dump with
bad value origin tracking~\cite{badapples} does not help with this bug.  In the
failing execution, origin tracking shows that the unusable lock is
uninitialized, but in the non-failing exeuction, origin tracking does not
monitor the {\em usable} lock value.  By contrast, last writer slicing tracks
that crucial information, helping debug this complex concurrency bug.

@@

Bad Apple Tracking~\cite{badapples} is also a form of debugging with data
provenance information. This work records the {\em origin} of an unusable value
and can report it to a programmer when it is used in a way that causes a
failure.
